---
title: "Signal-based Asset Pricing: Replicating the Change in Analyst Recommendation Factor"
subtitle: "Empirical Methods 2026 — Seminar Paper"
author: "Lukas Lobnig"
date: "2026-03-31"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    theme: cosmo
    fig-width: 8
    fig-height: 5
  pdf:
    toc: true
    number-sections: true
    fig-width: 7
    fig-height: 4.5
    geometry: margin=2.5cm
bibliography: references.bib
execute:
  warning: false
  message: false
---

# Introduction

This paper replicates and evaluates the **ChangeInRecommendation** signal from Chen and Zimmermann [-@ChenZimmermann2022], which originates from Jegadeesh, Kim, Krische, and Lee [-@JKKL2004]. The signal captures the quarterly change in sell-side analysts' consensus stock recommendation. JKKL show that while the *level* of analyst consensus loses predictive power after controlling for momentum and other characteristics, the *change* in consensus remains a robust cross-sectional return predictor that appears to contain information orthogonal to a large set of quantitative signals.

The economic intuition is straightforward: when analysts collectively revise their recommendations upward (downward), this reflects new private or semi-private information about a firm's prospects that has not yet been fully impounded into prices. Unlike the level of the recommendation --- which JKKL show is heavily contaminated by a "glamour bias" driven by sell-side incentive structures --- changes strip out this persistent bias and isolate the incremental information content of analyst research.

This paper proceeds as follows. Section 2 describes the data. Section 3 constructs the long-short factor portfolio and benchmarks it against Table 2 of Chen and Zimmermann [-@ChenZimmermann2022]. Section 4 tests the factor's pricing power against CAPM, FF3, FF-C4, and FF5. Section 5 evaluates whether augmenting existing models with the new factor improves their ability to price test portfolios. Section 6 examines factor competition. Section 7 presents cross-sectional Fama-MacBeth regressions, and Section 8 concludes.

```{r setup}
#| label: setup

suppressPackageStartupMessages({
  library(tidyverse)
  library(RSQLite)
  library(dbplyr)
  library(broom)
  library(lmtest)
  library(sandwich)
  library(ggplot2)
  library(kableExtra)
  library(GRS.test)
  library(scales)
})

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold"),
      legend.position = "bottom"
    )
)
```


# Data

## Database Connection and Signal Identification

We use the `tidy_finance.sqlite` database, which contains CRSP monthly stock data pre-merged with the Chen and Zimmermann [-@ChenZimmermann2022] signal characteristics.

```{r db-connect}
#| label: db-connect

db_path <- "/home/shared/data/tidy_finance.sqlite"
tidy_finance <- dbConnect(SQLite(), db_path, extended_types = TRUE)
```

```{r find-signal}
#| label: find-signal

# Identify the correct column for ChangeInRecommendation
all_cols <- tbl(tidy_finance, "crsp_monthly_signal") |> colnames()
signal_candidates <- grep("change|rec|analyst", all_cols, ignore.case = TRUE, value = TRUE)
cat("Candidate signal columns:\n")
print(signal_candidates)
```

```{r load-data}
#| label: load-data

# ── ADJUST THIS if the column name differs in your database ──
SIGNAL_COL <- "change_in_recommendation"

crsp_signal <- tbl(tidy_finance, "crsp_monthly_signal") |>
  select(
    permno, gvkey, month, ret_excess, ret, mktcap, mktcap_lag,
    exchange, all_of(SIGNAL_COL)
  ) |>
  collect() |>
  rename(signal = all_of(SIGNAL_COL)) |>
  # ── Timing fix: lag signal by one month ──────────────────────────────────
  # In crsp_monthly_signal, signal at month t is computed using data through
  # the END of month t (C&Z convention: "data at end of t can be used to trade
  # at end of t"). In CRSP, ret_excess at month t is the return DURING month t
  # (end of t-1 → end of t). To avoid look-ahead bias, we must use signal(t-1)
  # to predict ret_excess(t), i.e. lag the signal by one month.
  # This is consistent with the Fama-MacBeth section which uses lead(ret_excess).
  arrange(permno, month) |>
  group_by(permno) |>
  mutate(signal = lag(signal)) |>
  ungroup() |>
  drop_na(signal, ret_excess, mktcap, mktcap_lag) |>
  filter(mktcap_lag > 0)

# ── Load factor data ──
# Factor tables use "date"; we rename to "month" for consistency.
ff3 <- tbl(tidy_finance, "factors_ff3_monthly") |> collect() |> rename(month = date)
ff5 <- tbl(tidy_finance, "factors_ff5_monthly") |> collect() |> rename(month = date)
mom <- tbl(tidy_finance, "factors_momentum_monthly") |> collect() |> rename(month = date)

# Merge: FF3 + Momentum → FF-C4; add RMW/CMA from FF5 for FF5 tests.
# Use explicit select() to avoid duplicate columns (FF3 and FF5 share smb, hml, etc.)
factors <- ff3 |>
  select(month, risk_free, mkt_excess, smb, hml) |>
  inner_join(mom |> select(month, mom), by = "month") |>
  inner_join(ff5 |> select(month, rmw, cma), by = "month")

cat(
  "Sample period:", as.character(min(crsp_signal$month)), "to",
  as.character(max(crsp_signal$month)), "\n",
  "Stock-month observations:", scales::comma(nrow(crsp_signal)), "\n",
  "Unique firms:", scales::comma(n_distinct(crsp_signal$permno)), "\n",
  "Factor months:", nrow(factors)
)
```

## Descriptive Statistics

The ChangeInRecommendation signal is discrete and lumpy: many stocks experience no change in consensus recommendations in a given month, producing a large mass at zero. JKKL [-@JKKL2004, Table II Panel B] explicitly handle this by placing all "no change" stocks in the middle quintile, resulting in unequal-sized portfolios. This feature is central to understanding the sorting methodology.

```{r desc-stats}
#| label: desc-stats

coverage <- crsp_signal |>
  group_by(month) |>
  summarise(
    n_firms = n(),
    mean_signal = mean(signal),
    pct_zero = mean(abs(signal) < 1e-8),
    .groups = "drop"
  )

# ── Diagnostic: warn if no-change stocks are missing ─────────────────────
# JKKL explicitly place all "no-change" (signal = 0) stocks in the middle
# quintile, producing ~30% of the sample there. If pct_zero is near 0,
# the database stores NA instead of 0 for no-change stocks, and our drop_na
# has removed them — causing inflated extreme-quintile returns.
cat(sprintf(
  "Mean fraction of zero-signal obs: %.1f%%\n(Expected ~25-35%% for JKKL-style signal; near 0%% → NA-for-no-change issue)\n",
  mean(coverage$pct_zero) * 100
))

ggplot(coverage, aes(x = month, y = n_firms)) +
  geom_line(color = "steelblue") +
  labs(title = "Number of Firms with ChangeInRecommendation Signal",
       x = NULL, y = "Number of firms")
```

```{r desc-signal}
#| label: desc-signal

crsp_signal |>
  summarise(
    Mean = mean(signal), Median = median(signal), SD = sd(signal),
    P5 = quantile(signal, 0.05), P25 = quantile(signal, 0.25),
    P75 = quantile(signal, 0.75), P95 = quantile(signal, 0.95),
    `Frac = 0` = mean(abs(signal) < 1e-8)
  ) |>
  kable(digits = 4, caption = "Cross-sectional summary of the ChangeInRecommendation signal") |>
  kable_styling(bootstrap_options = "striped")
```

```{r signal-histogram}
#| label: signal-histogram

crsp_signal |>
  filter(signal >= quantile(signal, 0.01) & signal <= quantile(signal, 0.99)) |>
  ggplot(aes(x = signal)) +
  geom_histogram(bins = 80, fill = "steelblue", alpha = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Distribution of ChangeInRecommendation",
       subtitle = "Trimmed at 1st/99th percentiles; dashed red line at zero",
       x = "Signal value", y = "Count")
```


# Factor Construction and Replication

## Portfolio Sorting Methodology

The assignment requires us to "create the factor exactly as in the paper" and "use the signal to sort stocks into quantile portfolios (as given)" [@ChenZimmermann2022, Table 2]. JKKL [-@JKKL2004] use **quintile** sorts (5 portfolios), report **equal-weighted** returns, and note that the middle quintile absorbs all "no change" firms, producing unequal-sized portfolios (Table II, Panel B). Chen and Zimmermann follow each original paper's sorting scheme for their Table 2 replication, but standardize the breakpoint computation to use **NYSE-listed stocks only** to prevent small-cap Nasdaq stocks from distorting the cutoffs.

We therefore implement the following:

1. **Primary specification (C&Z Table 2 match)**: Quintile sorts, NYSE breakpoints, equal-weighted returns --- consistent with both JKKL's design and C&Z's standardization.
2. **Robustness**: Value-weighted returns and decile sorts for comparison.

```{r assign-portfolios}
#| label: assign-portfolios

# ── Portfolio assignment: NYSE breakpoints, month-by-month ──
# We iterate over months to handle edge cases cleanly and avoid
# issues with list columns inside grouped mutate().

assign_portfolios <- function(data, n_groups = 5) {
  
  months_vec <- sort(unique(data$month))
  
  map_dfr(months_vec, function(m) {
    
    d <- data |> filter(month == m)
    
    # NYSE breakpoints
    nyse_signals <- d |> filter(exchange == "NYSE") |> pull(signal)
    
    if (length(nyse_signals) < n_groups * 2) return(NULL)
    
    # Compute breakpoints at quintile/decile boundaries from NYSE stocks
    probs <- seq(0, 1, length.out = n_groups + 1)
    bp <- quantile(nyse_signals, probs = probs, na.rm = TRUE)
    
    # Assign ALL stocks using NYSE breakpoints
    # all.inside = TRUE ensures min → group 1 and max → group n_groups
    d |> mutate(portfolio = findInterval(signal, bp, all.inside = TRUE))
  })
}

# Primary: quintiles (JKKL design)
crsp_q5 <- assign_portfolios(crsp_signal, n_groups = 5)

# Robustness: deciles
crsp_d10 <- assign_portfolios(crsp_signal, n_groups = 10)
```

```{r portfolio-diagnostics}
#| label: portfolio-diagnostics

# ── Diagnostic: average portfolio sizes ──
# Due to the lumpy signal, some portfolios may be much larger (especially 
# the middle quintile which absorbs zero-change stocks).
crsp_q5 |>
  group_by(portfolio) |>
  summarise(
    `Mean N per month` = n() / n_distinct(month),
    `Mean signal` = mean(signal),
    .groups = "drop"
  ) |>
  kable(digits = 2, caption = "Quintile portfolio composition (NYSE breakpoints)") |>
  kable_styling(bootstrap_options = "striped")
```

## Portfolio Returns

JKKL [-@JKKL2004] report **equal-weighted** returns as their primary specification (Tables III, IX). We compute both EW and VW returns, but use EW as the primary specification to match the original paper.

```{r portfolio-returns}
#| label: portfolio-returns

compute_port_returns <- function(data) {
  data |>
    group_by(month, portfolio) |>
    summarise(
      ret_ew = mean(ret_excess, na.rm = TRUE),
      ret_vw = weighted.mean(ret_excess, w = mktcap_lag, na.rm = TRUE),
      n_stocks = n(),
      .groups = "drop"
    )
}

port_q5  <- compute_port_returns(crsp_q5)
port_d10 <- compute_port_returns(crsp_d10)
```

## Long-Short Factor Portfolio

The factor goes **long** the top quintile (stocks with the most positive recommendation changes, i.e. upgrades) and **short** the bottom quintile (most negative changes, i.e. downgrades). A positive mean return indicates upgrades outperform downgrades, consistent with JKKL's finding in Table III Panel C.

```{r long-short}
#| label: long-short

build_ls <- function(port_data, n_groups, weight_col = "ret_ew") {
  long  <- port_data |> filter(portfolio == n_groups) |>
    select(month, ret_long = all_of(weight_col))
  short <- port_data |> filter(portfolio == 1) |>
    select(month, ret_short = all_of(weight_col))
  
  inner_join(long, short, by = "month") |>
    mutate(ret_ls = ret_long - ret_short)
}

# Primary: quintile, EW (matches JKKL & C&Z Table 2)
ls_q5_ew  <- build_ls(port_q5, 5, "ret_ew")
# Robustness specifications
ls_q5_vw  <- build_ls(port_q5, 5, "ret_vw")
ls_d10_ew <- build_ls(port_d10, 10, "ret_ew")
ls_d10_vw <- build_ls(port_d10, 10, "ret_vw")
```

## Replication Results vs. Table 2

We compare the mean monthly return and $t$-statistic to the benchmark in Table 2 of Chen and Zimmermann [-@ChenZimmermann2022].

```{r table2-replication}
#| label: table2-replication

ls_summary <- function(ls_data, label) {
  r <- ls_data$ret_ls
  n <- length(r)
  tibble(
    Specification = label,
    `Mean (% p.m.)` = mean(r) * 100,
    `t-stat` = mean(r) / (sd(r) / sqrt(n)),
    `SD (% p.m.)` = sd(r) * 100,
    `Sharpe (ann.)` = mean(r) / sd(r) * sqrt(12),
    `N months` = n
  )
}

bind_rows(
  ls_summary(ls_q5_ew,  "Quintile EW (primary — JKKL style)"),
  ls_summary(ls_q5_vw,  "Quintile VW"),
  ls_summary(ls_d10_ew, "Decile EW"),
  ls_summary(ls_d10_vw, "Decile VW")
) |>
  kable(digits = 3, caption = "Long-short portfolio: replication summary") |>
  kable_styling(bootstrap_options = "striped")
```

```{r monotonicity-table}
#| label: monotonicity-table

port_q5 |>
  group_by(portfolio) |>
  summarise(
    `Mean EW (% p.m.)` = mean(ret_ew) * 100,
    `t (EW)` = mean(ret_ew) / (sd(ret_ew) / sqrt(n())),
    `Mean VW (% p.m.)` = mean(ret_vw) * 100,
    `t (VW)` = mean(ret_vw) / (sd(ret_vw) / sqrt(n())),
    `Mean N` = mean(n_stocks),
    .groups = "drop"
  ) |>
  kable(digits = 3,
        caption = "Quintile portfolio returns (1 = downgrades, 5 = upgrades)") |>
  kable_styling(bootstrap_options = "striped")
```

```{r monotonicity-plot}
#| label: monotonicity-plot

port_q5 |>
  group_by(portfolio) |>
  summarise(mean_ret = mean(ret_ew) * 100, .groups = "drop") |>
  ggplot(aes(x = factor(portfolio), y = mean_ret)) +
  geom_col(fill = "steelblue", alpha = 0.85) +
  labs(title = "Mean Monthly Excess Returns by Signal Quintile",
       subtitle = "Equal-weighted, NYSE breakpoints (primary specification)",
       x = "Quintile (1 = Downgrade, 5 = Upgrade)",
       y = "Mean excess return (% p.m.)")
```

### Discussion of Replication Quality

**Timing convention.** The single most important methodological choice is the lag structure. In the `crsp_monthly_signal` database, the C&Z signals are computed using data through the end of month $t$ and stored at month $t$. CRSP's `ret_excess` for month $t$ covers the period *from* end of $t-1$ *to* end of $t$. Using `signal(t)` to predict `ret_excess(t)` would create look-ahead bias (analyst upgrades correlate contemporaneously with high returns because the market reacts to the news). We therefore lag the signal by one month before sorting, so that `signal(t-1)` — known before month $t$ begins — predicts `ret_excess(t)`. This aligns with JKKL's design, where portfolios are formed at quarter-end $t$ and hold positions starting in quarter $t+1$.

**Data source and sample period.** JKKL (2004) use Zacks analyst recommendation data covering 1985–1998. C&Z note explicitly that "Zacks data goes back much further than the easily accessible IBES recommendations on WRDS" and therefore deviate from the original by using IBES. The `tidy_finance` database likewise relies on IBES, which starts in 1993/1994. Our sample therefore runs from roughly **1994 to present** — not 1985–1998. This has two consequences: (1) our mean return should be compared to C&Z's IBES-based reproduction (their Table 2), not to JKKL's Zacks-based numbers directly; and (2) the extended out-of-sample period after the 2004 publication date is subject to post-publication decay [@McLeanPontiff2016].

**Other construction differences.** C&Z typically winsorize signals at the 1st/99th percentiles before sorting; the pre-computed signal in the database may apply different or no winsorization. The discrete nature of recommendation changes also means that small differences in breakpoint computation --- such as how tied values at zero are allocated --- can shift the composition of extreme portfolios.


# Pricing Tests

We evaluate whether the long-short factor earns a statistically significant alpha after adjusting for exposure to established risk factors. A significant alpha means the signal captures a return pattern not explained by the benchmark model. We use Newey-West [-@NeweyWest1987] standard errors with 6 lags.

```{r pricing-tests}
#| label: pricing-tests

# Merge primary L/S factor with pricing factors
factor_data <- ls_q5_ew |>
  inner_join(factors, by = "month") |>
  drop_na()

run_alpha_test <- function(data, formula_str, model_name) {
  fit <- lm(as.formula(formula_str), data = data)
  ct <- coeftest(fit, vcov = NeweyWest(fit, lag = 6))
  tibble(
    Model = model_name,
    `Alpha (% p.m.)` = ct["(Intercept)", "Estimate"] * 100,
    `t-stat (NW)` = ct["(Intercept)", "t value"],
    `R²` = summary(fit)$r.squared
  )
}

bind_rows(
  run_alpha_test(factor_data, "ret_ls ~ mkt_excess", "CAPM"),
  run_alpha_test(factor_data, "ret_ls ~ mkt_excess + smb + hml", "FF3"),
  run_alpha_test(factor_data, "ret_ls ~ mkt_excess + smb + hml + mom", "FF-C4"),
  run_alpha_test(factor_data, "ret_ls ~ mkt_excess + smb + hml + rmw + cma", "FF5")
) |>
  kable(digits = 3,
        caption = "Alpha of the ChangeInRecommendation factor (NW lag=6)") |>
  kable_styling(bootstrap_options = "striped")
```

```{r full-c4-regression}
#| label: full-c4-regression

# Full FF-C4 regression — the key benchmark given JKKL's focus on momentum
fit_c4 <- lm(ret_ls ~ mkt_excess + smb + hml + mom, data = factor_data)
ct_c4 <- coeftest(fit_c4, vcov = NeweyWest(fit_c4, lag = 6))

tidy(ct_c4) |>
  mutate(
    term = recode(term, "(Intercept)" = "Alpha"),
    estimate = ifelse(term == "Alpha", estimate * 100, estimate)
  ) |>
  rename(Coefficient = term, Estimate = estimate,
         `SE (NW)` = std.error, `t` = statistic, `p` = p.value) |>
  kable(digits = 4, caption = "FF-C4 regression (alpha in % p.m.)") |>
  kable_styling(bootstrap_options = "striped")

cat("R²:", round(summary(fit_c4)$r.squared, 4))
```

The FF-C4 test is the most important benchmark. JKKL [-@JKKL2004, Table VIII Panel B] show that CHGCON remains significant even after controlling for momentum, contrarian scores, and all 12 individual quantitative signals. If our alpha survives the Carhart model --- which includes the UMD momentum factor --- this corroborates JKKL's core finding that recommendation changes convey genuinely new information beyond mechanical price signals.


# Factor Augmentation

## GRS Test on Industry Portfolios

We test whether adding the ChangeInRecommendation factor improves the FF-C4 model's ability to jointly price 30 industry portfolios, using the Gibbons-Ross-Shanken [-@GibbonsRossShanken1989] test.

```{r grs-test}
#| label: grs-test
# 1. Pull the data (already in wide format)
industries_30 <- tbl(tidy_finance, "industries_30_ff_monthly") |> collect()

# 2. Skip pivot_wider, just rename 'date' to 'month' to match your factors data, and drop NAs
ind_wide <- industries_30 |>
  rename(month = date) |> 
  drop_na()

# 3. Define your industry columns (everything except the 'month' column)
ind_cols <- setdiff(colnames(ind_wide), "month")

# 4. Pull your custom factor
chgrec_factor <- ls_q5_ew |> select(month, chgrec = ret_ls)

# 5. Join everything together
test_data <- ind_wide |>
  inner_join(factors, by = "month") |>
  inner_join(chgrec_factor, by = "month") |>
  drop_na()

# 6. Set up the matrices for the GRS test
Y  <- test_data |> select(all_of(ind_cols)) |> as.matrix()
X1 <- test_data |> select(mkt_excess, smb, hml, mom) |> as.matrix()
X2 <- test_data |> select(mkt_excess, smb, hml, mom, chgrec) |> as.matrix()

# 7. Run the tests
grs1 <- GRS.test(Y, X1)
grs2 <- GRS.test(Y, X2)

# 8. Output the results
tibble(
  Model = c("FF-C4", "FF-C4 + ChgRec"),
  `GRS stat` = c(grs1$GRS.stat, grs2$GRS.stat),
  `p-value` = c(grs1$GRS.pval, grs2$GRS.pval)
) |>
  kable(digits = 4, caption = "GRS test on 30 industry portfolios") |>
  kable_styling(bootstrap_options = "striped")
```

## Individual Alpha Comparison

```{r spanning-alphas}
#| label: spanning-alphas

compute_alphas <- function(test_data, ind_cols, rhs) {
  map_dfr(ind_cols, function(col) {
    df <- test_data |> rename(y = all_of(col))
    fit <- lm(as.formula(paste("y ~", rhs)), data = df)
    ct <- coeftest(fit, vcov = NeweyWest(fit, lag = 6))
    tibble(portfolio = col,
           alpha = ct["(Intercept)", "Estimate"] * 100,
           t_stat = ct["(Intercept)", "t value"])
  })
}

a1 <- compute_alphas(test_data, ind_cols, "mkt_excess + smb + hml + mom") |>
  rename(alpha_c4 = alpha, t_c4 = t_stat)
a2 <- compute_alphas(test_data, ind_cols, "mkt_excess + smb + hml + mom + chgrec") |>
  rename(alpha_aug = alpha, t_aug = t_stat)

left_join(a1, a2, by = "portfolio") |>
  mutate(reduction = abs(alpha_c4) - abs(alpha_aug)) |>
  summarise(
    `Mean |α| FF-C4 (%)` = mean(abs(alpha_c4)),
    `Mean |α| Augmented (%)` = mean(abs(alpha_aug)),
    `Mean reduction (pp)` = mean(reduction),
    `# improved` = sum(reduction > 0),
    `Total` = n()
  ) |>
  kable(digits = 3, caption = "Alpha reductions from augmentation") |>
  kable_styling(bootstrap_options = "striped")
```


# Factor Competition

```{r spanning-test}
#| label: spanning-test

# ChgRec regressed on FF-C4 factors → is the new factor spanned?
fit_span <- lm(chgrec ~ mkt_excess + smb + hml + mom, data = test_data)
ct_span <- coeftest(fit_span, vcov = NeweyWest(fit_span, lag = 6))

tidy(ct_span) |>
  mutate(estimate = ifelse(term == "(Intercept)", estimate * 100, estimate)) |>
  kable(digits = 4,
        caption = "Spanning: ChgRec on FF-C4 (intercept in % p.m.)") |>
  kable_styling(bootstrap_options = "striped")

cat("R²:", round(summary(fit_span)$r.squared, 4))
```

```{r reverse-regressions}
#| label: reverse-regressions

map_dfr(c("mkt_excess", "smb", "hml", "mom"), function(fac) {
  df <- test_data |> rename(y = all_of(fac))
  fit <- lm(y ~ chgrec, data = df)
  ct <- coeftest(fit, vcov = NeweyWest(fit, lag = 6))
  tibble(`Dep. Var` = fac,
         `α (% p.m.)` = ct[1,1]*100, `t(α)` = ct[1,3],
         `β(ChgRec)` = ct[2,1], `t(β)` = ct[2,3],
         `R²` = summary(fit)$r.squared)
}) |>
  kable(digits = 4, caption = "Reverse: each FF-C4 factor on ChgRec") |>
  kable_styling(bootstrap_options = "striped")
```

```{r correlation-matrix}
#| label: correlation-matrix

test_data |>
  select(mkt_excess, smb, hml, mom, chgrec) |>
  cor() |> round(3) |>
  kable(caption = "Factor correlation matrix") |>
  kable_styling(bootstrap_options = "striped")
```

A low $R^2$ combined with a significant intercept is the clearest evidence the factor is not redundant. This aligns with JKKL's finding that recommendation changes carry information "orthogonal to a large range of other predictive variables" [-@JKKL2004].


# Comparison to Table 2 and Literature

```{r cumulative-returns}
#| label: cumulative-returns

ls_q5_ew |>
  arrange(month) |>
  mutate(cum_ret = cumprod(1 + ret_ls) - 1) |>
  ggplot(aes(x = month, y = cum_ret)) +
  geom_line(color = "steelblue", linewidth = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  scale_y_continuous(labels = percent) +
  labs(title = "Cumulative Returns: ChangeInRecommendation Long-Short Factor",
       subtitle = "Quintile 5 – Quintile 1, equal-weighted",
       x = NULL, y = "Cumulative excess return")
```

Potential sources of deviation from C&Z Table 2:

- **Sample period**: The `tidy_finance` database may cover a different window than C&Z's full extract.
- **Signal construction**: C&Z compute the signal from raw IBES using their own code. Differences in how the recommendation change is computed (lag structure, analyst filtering) affect outcomes.
- **Breakpoint sensitivity**: With a lumpy signal, small differences in how tied values at zero are handled shift the extreme portfolio boundaries.
- **Post-publication decay**: JKKL published in 2004. McLean and Pontiff [-@McLeanPontiff2016] document that anomalies weaken by ~58% post-publication on average.


# Cross-Sectional Pricing (Voluntary)

## Fama-MacBeth Regressions

Following Fama and MacBeth [-@FamaMacBeth1973], we run monthly cross-sectional regressions of next-month excess returns on the signal. We use Newey-West standard errors with 3 lags and signal *characteristics* (not factor betas) as the regressor, consistent with the assignment instructions and with JKKL's approach.

```{r fama-macbeth}
#| label: fama-macbeth

fm_data <- crsp_signal |>
  # signal is already lagged (see load-data); use ret_excess directly
  drop_na(ret_excess, signal)

fm_coefs <- fm_data |>
  group_by(month) |>
  do(tidy(lm(ret_excess ~ signal, data = .))) |>
  ungroup()

gamma1_ts <- fm_coefs |> filter(term == "signal")

fm_stat <- function(x, lag = 3) {
  mu <- mean(x)
  se <- sqrt(sandwich::lrvar(x, type = "Newey-West", lag = lag))
  c(mean = mu, se = se, t = mu / se)
}

g1 <- fm_stat(gamma1_ts$estimate)

tibble(
  `γ₁ (% p.m.)` = g1["mean"] * 100,
  `NW SE (%)` = g1["se"] * 100,
  `t-stat (NW, lag=3)` = g1["t"],
  `N months` = nrow(gamma1_ts)
) |>
  kable(digits = 3, caption = "Fama-MacBeth: univariate signal premium") |>
  kable_styling(bootstrap_options = "striped")
```

## Extended Regression with Size Control

```{r fm-with-size}
#| label: fm-with-size

fm_data_ext <- fm_data |>
  mutate(log_size = log(mktcap_lag)) |> drop_na(log_size)

fm_coefs_ext <- fm_data_ext |>
  group_by(month) |>
  do(tidy(lm(ret_excess ~ signal + log_size, data = .))) |>
  ungroup()

g1x <- fm_stat(fm_coefs_ext |> filter(term == "signal") |> pull(estimate))
gs  <- fm_stat(fm_coefs_ext |> filter(term == "log_size") |> pull(estimate))

tibble(
  Variable = c("Signal (γ₁)", "Log size"),
  `Mean (% p.m.)` = c(g1x["mean"], gs["mean"]) * 100,
  `NW t-stat` = c(g1x["t"], gs["t"])
) |>
  kable(digits = 3, caption = "Fama-MacBeth with size control") |>
  kable_styling(bootstrap_options = "striped")
```

## Dynamics of the Premium

```{r gamma-dynamics}
#| label: gamma-dynamics

gamma1_ts |>
  mutate(
    gamma_pct = estimate * 100,
    gamma_roll = zoo::rollmean(gamma_pct, k = 36, fill = NA, align = "right")
  ) |>
  ggplot(aes(x = month)) +
  geom_line(aes(y = gamma_pct), alpha = 0.2, color = "grey50") +
  geom_line(aes(y = gamma_roll), color = "steelblue", linewidth = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = as.Date("2004-06-01"), linetype = "dotted", color = "red") +
  annotate("text", x = as.Date("2005-06-01"),
           y = max(gamma1_ts$estimate * 80, na.rm = TRUE),
           label = "JKKL publication", hjust = 0, color = "red", size = 3) +
  labs(title = "Time-Varying Signal Premium (γ₁)",
       subtitle = "Monthly (grey) and 36-month rolling mean (blue)",
       x = NULL, y = expression(hat(gamma)[1] ~ "(% p.m.)"))
```

```{r subsample}
#| label: subsample

gamma1_ts |>
  mutate(period = ifelse(month <= as.Date("2004-06-01"),
                         "Pre-publication (≤2004)", "Post-publication (>2004)")) |>
  group_by(period) |>
  summarise(
    `Mean γ₁ (% p.m.)` = mean(estimate) * 100,
    `SD (%)` = sd(estimate) * 100,
    `t-stat` = mean(estimate) / (sd(estimate) / sqrt(n())),
    `N months` = n(), .groups = "drop"
  ) |>
  kable(digits = 3, caption = "Signal premium: pre- vs. post-publication") |>
  kable_styling(bootstrap_options = "striped")
```


# Conclusion

This paper replicates the ChangeInRecommendation factor from Chen and Zimmermann [-@ChenZimmermann2022], based on the quarterly change in consensus analyst recommendations studied by JKKL [-@JKKL2004].

[Fill in based on your results: Report the mean return and t-stat vs. Table 2. Discuss whether the FF-C4 alpha is significant. Report factor competition R² and correlation with MOM. Note GRS improvement. Comment on Fama-MacBeth premium dynamics and post-publication decay.]

From a broader perspective, JKKL's distinction between the *level* and the *change* in consensus recommendations is a powerful methodological insight. The level absorbs sell-side incentive biases --- the preference for glamour stocks --- that erode its predictive content, while the change strips out this persistent component and isolates incremental news. This mirrors the logic of momentum as a change in prices, earnings surprises as changes in expectations, and forecast revisions as changes in beliefs. The general lesson is that for any signal shaped by persistent institutional biases, its first difference may be more informative than its level.


# References {.unnumbered}

::: {#refs}
:::

```{r cleanup}
#| label: cleanup
#| include: false
dbDisconnect(tidy_finance)
```
